Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added two references on circular trees; revision of section 1 and introductions of sections 3 and 4
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2877 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 59239c84ae2c865e9bf371494133fec4fbb77d28

In many approaches to modeling circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.

One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter, to be modeled not only in separate models but also in a 
mixed model. There, within an iterative optimization algorithm the two parameters are estimated and 
improved stepwise until convergence. To obtain proper initial values it is recommended to apply
two separate models where either one of the parameters is fixed while the other is modeled depending
on covariates. While being built on well elaborated theory this methodology and the contained 
optimization problem have shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function that is not globally concave, 
and thus bear a high risk of not converging if not applied thoughtfully 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}.
An alternative approach to regression modeling of circular data that avoids this strong dependence 
on choosing well-fitting initial values is presented by \cite{Mulder+Klugkist:2017} employing a GLM 
in a Bayesian framework.

\fixme{revise literature survey with focus on transition to GLM}

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they provide a powerful toolbox for modeling circular data, particularly in case of
smooth effects, they may not be able to capture all types of effects. For example, abrupt changes 
might not be detected properly owing to the enforced additive structure and model assumptions 
such as the type of link function and the form of linear combinations.

\fixme{revise transition to trees/forests}

Therefore, we propose a very flexible alternative approach by adapting regression trees and random 
forests to circular data, specifically by employing the von Mises distribution within the methodology
of distributional trees and distributional forests following \cite{Schlosser+Hothorn+Stauffer:2019}. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

\fixme{include and revise ``motivating'' example}

As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the
following hour.  Innsbruck lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.

@@

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

A more extensive case study of circular regression trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna is presented in Section~\ref{sec:wind} where their performances are
benchmarked to alternative commonly used approaches.  Beforehand,
distributional models for circular data are discussed in
Section~\ref{sec:prob_circ} and the methodology of circular regression trees
and forest and its features are presented in Section~\ref{sec:tree_forest}.
After the case study, a comprehensive summary and conlusion is given in
Section~\ref{sec:summary}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on the covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs \citep{Fisher+Lee:1992, Fisher:1993}.
In a more general setting this idea can be extended by generalized additive models for location, scale, 
and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} where each distribution parameter is estimated
by its own generalized additive model (GAM). However, an adaption of this methodology to circular data
can not be obtained by the straight forward approach of simply wrapping the modelled additive effects
around the unit circle since the parameters of the von Mises distribution and their dependence/interaction 
differs strongly from the linear case and therefore requires specific handling.
Moreover, difficulties might arise in case of non-additive changes and specifying a model properly can 
be very challenging, particularly for a high number of covariates and no information on possible 
interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

\fixme{information on already available approaches to wind direction forecasting}

Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
area, is considered in this study.
In that way completely different geographic features and their possible influences
on wind direction are included which allows to test for a high flexibility of the applied methods.
For the two locations the above introduced methodology of circular distributional trees and forests is
employed to wind direction data and additional meteorological measurements in order to obtain hourly 
wind direction forecasts, separately for Innsbruck and Vienna.

@@

Lisa Schlosser recieved a PhD scholarship granted from the University of Innsbruck.

>>>>

In many approaches to modeling circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. There, within an iterative 
optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
To obtain proper initial values it is recommended to apply two separate models where either one of 
the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology and the contained 
optimization problem have shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, the selection of 
starting values is crucial and bears a high risk of the model not to converge if not done thoughtfully. 
In order to avoid this strong dependence on choosing well-fitting initial values \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
in a Bayesian framework.

%\fixme{revise literature survey with focus on transition to GLM}

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they generally provide a powerful toolbox for modeling particularly smooth effects, 
it has to be considered that representing these smooth effects on the unit circle
can be challenging. Moreover, the discussed additive models may not be able to capture all 
types of effects adequately. For example, abrupt changes might not be detected properly owing 
to the enforced additive structure and model assumptions such as the type of link 
function and the form of linear combinations.

An alternative modeling approach that can deal with non-additive effects is provided
by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
which was further explored by \cite{Larraondo+Inza+Lozano:2018}. This methodology
enables splits leading to subsets which contain the ``origin point'' $0$ equal to $2\pi$
and in that way allows for a proper representation of periodicity.

Following the idea of employing tree-based methods but including also the probabilistic concept
of regressing distribution parameters on covariates we propose a very flexible approach 
by applying the von Mises distribution within the methodology of distributional trees and 
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional tree applied to 
predict wind directions at the airport of Innsbruck in the following hour. Wind direction is 
a classical circular quantity and accurate forecasts are of great importance for decision-making 
processes and risk management in air traffic management. Innsbruck lies at the bottom of a 
deep valley in the Alps where topography channels wind along the west-east valley axis or along 
a tributary valley intersecting from the south. Hence, pressure gradients to which valley wind 
regimes react are considered as covariates along with other meteorological measurements at the 
airport (lagged by one hour), such as wind direction and wind speed at the airport.

@@

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting 
at the airport of Innsbruck for the following hour.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

A more extensive case study of circular regression trees and forests applied to 
probabilistic forecasting of wind direction at the airports of Innsbruck and Vienna 
is presented in Section~\ref{sec:wind} where their performances are benchmarked to 
alternative commonly used approaches.  Beforehand, distributional models for circular 
data are discussed in Section~\ref{sec:prob_circ} and the methodology of circular 
regression trees and forest and its features are presented in Section~\ref{sec:tree_forest}.
After the case study, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. While applying the mixed model 
requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
However, an adaption of this methodology to circular data can be very complex due to a
challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
Moreover, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very tough task, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
and also supported by the results illustrated in the terminal nodes of the tree in 
Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
mountain chains are a strongly influential component of the behavior of wind directions.
While this provides an interesting and challenging environment for testing different
statistical models it has to be considered as a rather unusual location for an airport.
Therefore, in order to allow for a broad and more general evaluation and also to test
for a high flexibility of the applied methods a second site with quite converse geographic 
features is included in this study as well.
Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
area, is considered.
For the two locations the above introduced methodology of circular distributional trees and forests is
employed to wind direction data and additional meteorological measurements in order to obtain hourly 
wind direction forecasts, separately for Innsbruck and Vienna.

@@

Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.
